---
title: "SHINYREPS_PROJECT"
output:
  html_document:
    toc: yes
    css: styles.css
---

<div class="contentbox">

```{r Initialise, echo=F,result='hide',error=F,warning=F,message=F}
source('ChIP.shinyrep.helpers.R')
loadGlobalVars()
targets <<- ChIPhelper.init("readTargets")
peaks   <<- ChIPhelper.init("readPeaks")
``` 


## Raw reads QC ##

The raw sequence reads of all samples are analysed with the popular FastQC tool (http://www.bioinformatics.babraham.ac.uk/projects/fastqc/).

1. The "Duplication levels" plots show the proportion of the library which is made up of sequences in each of the different duplication level bins. It is useful for the assessment of the NGS library complexity and the presence or absence of PCR over-amplification during library prep. There are two lines on the plot. The blue line takes the full sequence set and shows how its duplication levels are distributed. In the red plot the sequences are de-duplicated and the proportions shown are the proportions of the deduplicated set which come from different duplication levels in the original data. In a properly diverse library most sequences should fall into the far left of the plot in both the red and blue lines. A general level of enrichment, indicating broad oversequencing in the library will tend to flatten the lines, lowering the low end and generally raising other categories. The presence of dominant sequences (e.g. adapter dimers) will tend to produce spikes towards the right of the plot. These high duplication peaks will most often appear in the red trace as they make up a high proportion of the original library, but usually disappear in the blue trace as they make up an insignificant proportion of the deduplicated set. If peaks persist in the blue trace then this suggests that there are a large number of different highly duplicated sequences which might indicate either a contaminant set or a severe PCR duplication. See the related PBC metrics below.

2. The "Read qualities" Box-Whisker plots show the range of quality values across all base positions:
    (i) The central red line is the median value,
    (ii) The yellow box represents the inter-quartile range (25-75%),
    (iii) The upper and lower whiskers represent the 10% and 90% points,
    (iv) The blue line represents the mean quality.
The y-axis on the graph shows the Phred quality scores, which are logarithmically related to the base-calling error probabilities. The higher the score the better the base call. The background of the graph divides the y axis into very good quality calls (green), calls of reasonable quality (orange), and calls of poor quality (red). Typically, the majority of calls on all base positions fall into the green area.

3. The "Sequence bias" plots show the proportion of each base (% G, A, T and C) at each position. In a typical ChIP-seq library there would be little difference between the positions of a sequence run, so the lines in this plot should run parallel with each other, G being similar to C and A to T. 

```{r FastQC_paragraph, echo=F,results='asis',error=F,warning=F,message=F}
cat(ChIPhelper.Fastqc(web=F),sep="\n")
```

## Read mapping ##

Mapping to the reference genome was done using the established NGS mapping tool Bowtie2 (http://bowtie-bio.sourceforge.net/bowtie2). The mapping statistics show the total number of reads delivered to the aligner ("all reads"), the percentage of paired-end (PE) reads ("paired"), the number of PE reads which are failed to align concordantly ("unmapped"), the number of PE reads which have a unique alignment ("uniquely mapped") and the number of PE reads which are mapped to multiple places ("multiple mapped"):

```{r Bowtie_paragraph, echo=F,results='asis',error=F,warning=F,message=F}
cat(ChIPhelper.Bowtie2PE(),sep="\n")
```

## Insert Size ##

```{r echo=F,results='asis',error=F,warning=F,message=F}

cat(DEhelper.insertsize(), sep="\n")

```

## PCR bottleneck coefficient ##

The PBC (PCR bottleneck coefficient) is an approximate measure of library complexity. Provisionally, 0-0.5 is severe bottlenecking, 0.5-0.8 is moderate bottlenecking, 0.8-0.9 is mild bottlenecking, while 0.9-1.0 is no bottlenecking.

Very low values can indicate a technical problem, such as PCR bias, or a biological finding, such as a very rare genomic feature. Nuclease-based assays (DNase, MNase) detecting features with base-pair resolution (transcription factor footprints, positioned nucleosomes) are expected to recover the same read multiple times, resulting in a lower PBC score for these assays. Note that the most complex library, random DNA, would approach 1.0, thus the very highest values can indicate technical problems with libraries. Some common numbers from ENCODE datasets are: 82% for TF ChIP, 89% for His ChIP, 77% for DNase, 98% for FAIRE, and 97% for control ENCODE datasets have no or mild bottlenecking.

```{r PBC_paragraph, echo=F,results='asis',error=F,warning=F,message=F}
cat(ChIPhelper.PBC(),sep="\n")
```

## IPstrength ##

To estimate the IP strength, we attempt to decompose the population of IP reads into two distinct components: those pulled down by the antibody, and background.

```{r IPstrength_paragraph, echo=F,results='asis',error=F,warning=F,message=F}
cat(ChIPhelper.IPstrength(web=F),sep="\n")
```

The plot is useful to detect several forms of biases:

* The distance between the two curves depends on the enrichment of the IP versus the input; close curves suggest weak or diffuse enrichment.
* Under-sequencing produces curves close to the x-axis for a large percentage of bins since no reads were aligned on this bins. 
* Sequencing bias towards a small set of specific regions produces curves shifted towards the y-axis for a large percentage of bins.


## ChIP-seq peaks ##

The number of peaks called at the default fasle-disovery rate (FDR) cutoff of 1% by MACS2 (https://github.com/taoliu/MACS/) in the different comparisons is shown below.

```{r PeakCall_paragraph, echo=F,results='asis',error=F,warning=F,message=F}
peaks <- ChIPhelper.init("readPeaks")

l <- lapply(peaks,function(x) {
	x <- table(x$chr)
	xx <- as.numeric(x)
	names(xx) <- names(x)
	xx
})
p <- Reduce(function(x,y) {
	z <- merge(x,y,all=T,by=0)
	rownames(z) <- z[,1]
	z[,-1]
},l)
#colnames(p) <- 1:ncol(p)
colnames(p) <- names(l)

# print the legend
#cat("",fill=T)
#cat(kable(data.frame(column=1:length(l),sample=names(l))),sep="\n")
#cat("",fill=T)

# print the peaks table
cat("",fill=T)
cat(kable(p,row.names=T),sep="\n")
cat("",fill=T)
```

## Correlation of peaks ##

The following Venn diagrams correlate the peaks called in the sample groups.

```{r peak_venn_paragraph, echo=F,results='asis',error=F,warning=F,message=F}
ChIPhelper.VennDiagram()
```

## Functional annotation of peaks ##

The following plot show the feature distribution based on their chromosome region 

### Feature distribution of peaks for all samples ###

```{r feature_distribution_paragraph, echo=F, results='asis', error=F, warning=F, message=F}
functionalPlot <- paste0( SHINYREPS_PEAK_ANNOTATION, "/", "ChIPseq_Feature_Distribution_Barplot.png")
cat("![Peak_Annotation img](", functionalPlot, ")",fill=T)
```

### Distribution of TF-binding loci relative to TSS for all samples ### 

```{r feature_distribution_TSS_paragraph, echo=F, results='asis', error=F, warning=F, message=F}
    functionalPlotTSS <- paste0( SHINYREPS_PEAK_ANNOTATION, "/", "ChIPseq_Feature_Distribution_Related_to_TSS_Barplot.png")
    cat("![Peak_Annotation img](", functionalPlotTSS, ")",fill=T)
```

### ChIP peaks coverage over Chromosomes ###  

```{r peak_annotation_coverage_paragraph, echo=F, results='asis', error=F, warning=F, message=F}
    cat(ChIPhelper.peakAnnotationCoverage(web=F),sep="\n")
```

### UpSet Plot ###

```{r peak_annotation_upset_paragraph, echo=F, results='asis', error=F, warning=F, message=F}
    cat(ChIPhelper.peakAnnotationUpSet(web=F),sep="\n")
```

## Used tools and versions for this analysis ##
```{r ToolVersions_paragraph, echo=F, results='asis', error=F, warning=F, message=F}
cat(Toolhelper.ToolVersions(), sep="\n")
```

</div>
