---
title: "SHINYREPS_PROJECT"
output:
  html_document:
    toc: yes
---

```{r echo=F,result='hide',error=F,warning=F,message=F}
source('variant.shinyrep.helpers.R')
loadGlobalVars()
# load(SHINYREPS_DE_EDGER,envir=.GlobalEnv)		# the outcome from the DE analysis
# VARhelper.init("prepareDEdataTable")
# VARhelper.init("renderUcscGeneLinks")
# VARhelper.init("prepareDistanceMatrix")
``` 

## Quality Control RNAseq ##

### Sequencing quality ###
The sequencing quality of the run was good, and the read distribution over the libraries was good.

### Raw reads qualities, sequence bias and duplication
```{r echo=F,results='asis',error=F,warning=F,message=F}
cat(VARhelper.Fastqc(web=F),sep="\n")
```

### Mapping
The mapping statistics show the total number of reads delivered to the aligner, the number of mapped reads and reads not mapping at all to the ref aligning to the wrong reference genome.

```{r echo=F,results='asis',error=F,warning=F,message=F}
cat(VARhelper.BWA(),sep="\n")
```

### Variant calling
Variant calling can be performed using GATK UnifiedGenotyper and GATK HaplotypeCaller.
Both tools include default read filters that exclude reads from variant calling.

#### GATK UnifiedGenotyper filtered read statistics
```{r echo=F,results='asis',error=F,warning=F,message=F}
cat(VARhelper.GATKug(),sep="\n")
```

#### GATK HaplotypeCaller filtered read statistics
```{r echo=F,results='asis',error=F,warning=F,message=F}
cat(VARhelper.GATKhc(),sep="\n")
```

#### GATK VariantEval variant statistics
Variant statistics show the basic properties of your variants as triplets for "all", "known" ,"novel" variants in comparison to dbSNP.
The calculated ratios should be similar in "all", "known" and "novel" subsets. Highly deviating values in the "novel" subset indicate a high false positive rate.
Due to selective pressure transitions are more likely to be retained in the genome than transversions.
For Exome sequencing a Ti/Tv ration of 3 would be good, while whole genome sequencing might result in values blow 2.


```{r echo=F,results='asis',error=F,warning=F,message=F}
cat(VARhelper.GATKvarianteval(),sep="\n")
```

## Used tools and versions for this analysis

```{r echo=F,results='asis',error=F,warning=F,message=F}
cat(VARhelper.ToolVersions(),sep="\n")
``` 
